### 2023-07-12 submission/manuscript.Rmd
Use gemelli to perform rclr/rpca

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
  - \usepackage[document]{ragged2e}
---

@@

${\dagger}$ To whom corresponsdence should be addressed:

@@

Considering it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences found across samples in a study, researchers need to control for the effect of uneven sequencing effort. How to do this has become a contentious question. Some have argued that rarefaction is "inadmissible" because it omits valid data. A number of alternative approaches have been developed to normalize and rescale the data that purport to be invariant to the number of observations. I generated community distributions based on 12 published datasets where I was able to assess the ability of multiple methods to control for uneven sequencing effort. Rarefaction was the only method that could control for variation in uneven sequencing effort when measuring commonly used alpha and beta diversity metrics. Next, I compared the false detection rate and power to detect true differences between simulated communities with a known effect size using various alpha and beta diversity metrics. Although all methods of controlling for uneven sequencing effort had an acceptable false detection rate when samples were randomly assigned to two treatment groups, rarefaction was consistently able to control for differences in sequencing effort when sequencing depth was confounded with treatment group. Finally, the statistical power to detect differences in alpha and beta diversity metrics was consistently the highest when using rarefaction. These simulations underscore the importance of using rarefaction to normalize the number of sequences across samples in amplicon sequencing analyses.

@@

The ability to generate millions sequence reads from 16S rRNA gene amplicons and metagenomic datasets has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that are embedded into the sequence of the PCR primers [@Sogin2006; @Caporaso2010; @Kozich2013]. Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1). This occurs because pooling of DNA from multiple PCRs is fraught with numerous opportunities for technical errors to compound leading to a skewed distribution. Aside from developing better methods of pooling DNA, the question of how to control for uneven sequencing effort in microbial ecology studies has become controversial.

The practice of rarefaction has been commonly used in ecology for more than 50 years as a tool to control for uneven sequencing effort across experimental replicates [@Sanders1968; @Hurlbert1971]. Microbial ecologists have used it to compare 16S rRNA gene sequence data for the past 25 years [@Dunbar1999; @Hughes2001; @Moyer1998]. With rarefaction the investigator selects a desired threshold of sequencing effort and removes any samples below that threshold. They then randomly select that many sequences, with replacement from each sample. Based on the observed sequence counts, the researcher can then calculate alpha diversity metrics including richness and diversity indices or beta diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index. I refer to this single sampling as a subsample; this method is implemented as the `sub.sample` function in mothur [@Schloss2009] and the `rrarefy` function in the `vegan` R package [@Dixon2003]. Rarefaction repeats the subsampling a large number of times (e.g., 100 or 1,000 times) and calculates the mean of the alpha or beta diversity metric over those subsamplings; rarefaction is implemented in mothur using the `summary.single` and `dist.shared` functions [@Schloss2009] and with the `vegan` R package using the `rarefy` or `avgdist` functions [@Dixon2003]. Rarefaction effectively tells a researcher what an alpha or beta diversity metric would have been for a collection of samples if they were all sequenced to the same depth. Although a closed form equation exists to calculate the expected richness [@Hurlbert1971], it is computationally easier to empirically calculate richness and other alpha and beta diversity metrics by rarefaction.

[@Schloss2023wnwn]
Confusion between rarefy and rarefaction
***In 2014, McMurdie and Holmes [@McMurdie2014] announced that rarefaction of microbial community data is "statistically inadmissible" because it omits valid data. In their simulations, they observed that the use of rarefaction reduced the statistical power to compare communities based on beta diversity metrics. The communities used in this study were simulated by mixing operational taxonomic unit (OTU) frequencies from marine and human fecal communities, which were then sampled to generate 40 replicate communities each sampled to different depths. The simulation with the greatest depth of coverage obtained an average of 10,000 sequences per sample. Analysis of the R code that they used indicates that the simulations were actually only based on a single subsample and not traditional rarefaction. Furthermore, the authors removed any OTU from the simulated communities that was not observed in three or more samples. This is a frequently used approach for analyzing microbiome data [@Bokulich2013]. Unfortunately, it removes valid data, skews the distribution of the community, and removes true patchiness from the representation of the communities [@Schloss2023sing]. Finally, the analysis did not include analysis of alpha diversity metrics and chose to assess beta diversity metrics based on clustering quality of samples rather than using the more frequently used approach of employing non-parametric multivariate analysis of variance [@McArdle2001]. Regardless of these caveats, McMurdie and Holmes recommended the use of variance stabilizing transformations such as those applied for gene expression analysis such as that found in the DESeq and DESeq2 R packages.***

**Since McMurdie and Holmes published their simulations, others have developed alternative approaches to controlling for uneven sequencing effort in amplicon sequencing studies. For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019]. Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [@Willis2015] that have been used in microbial ecology studies. For beta diversity metrics at least four approaches have been pursued. First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020]. Second, normalization strategies have been developed where the relative abundance is multiplied by the size of the minimum desired sequencing effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013]. Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014]. This strategy is purported to control for uneven sequencing effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021]. Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].**

The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the original study. My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1). The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects. For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sequencing effort. My overall conclusion was that rarefaction outperformed the alternative strategies.

@@

***Without rarefaction, metrics of beta diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring beta diversity to sequencing effort, I used the same null models used for studying the sensitivity of alpha diversity. Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3). First, I calculated the Jaccard distance coefficient between all pairs of communities within a dataset. The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs. Jaccard distances were calculated using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS. Only the distances calculated with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU. Similar to what I observed with the Jaccard coefficient, only the data normalized with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring OTUs in samples with zero counts (Robust CLR), adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR). The Aitchison distances were all strongly sensitive to sequencing effort (Figure 1B). Finally, I used the variance-stabilization transformation (VST) from DESeq2 prior to calculating Euclidean distances. Again, there was a strong sensitivity to sequencing effort (Figure 1B). Although Euclidean distances are not typically used on raw or rarefaction-normalized count data in ecology [@Legendre2012], Euclidean distances calculated with rarefaction were not sensitive to sequencing effort. Across each of the beta diversity metrics and approaches used to account for uneven sequencing effort and sparsity, rarefaction was the least sensitive approach to differences in sequencing effort.

@@

```{r coverage}
loss <- table_1 %>%
  select(nice_name, fold_difference) %>%
  mutate(loss = 100 - round(100* (1 - 1/fold_difference), 1)) %>%
  slice_max(loss)

@@

***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$loss`% of sequences that were in the largest sample). To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6). The richness values increased with sequencing effort as rare OTUs would continue be detected. In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sequencing effort. This result was expected since increased sequencing would lead to increased precision in the measured relative abundance of the OTUs. Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances. Although the mean richness appeared to increased unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable. In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths. Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase. Regardless, the coefficients of variation were relatively small. This analysis indicates that there are benefits to increased sequencing depths.

@@

All simulations have weaknesses and should be interpreted with caution. However, the simulated communities generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups. Indeed, is is perfectly admissible and proper to use rarefaction with microbial community data. The alternatie is to risk reaching unwarranted conclusions.

@@

**Null treatment models.** I created an unbiased and biased treatment model. In the unbiased model, samples were randomly assigned to one of two treatment groups. In the biased treatment model, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group. Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\upalpha$) of 0.05.

@@

**Richness-adjusted community model.** In the richness-adjusted community model, communities were randomly assigned to one of two simulated treatment groups. Communities in the first treatment group were generated by calculating the relative abundance of each OTU across all samples and using those values as the probability of sampling each OTU. This probability distribution was sampled until each sample had the same number of sequences that it did in the observed data. Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs. Sequences were sampled from the richness-adjusted community distribtion, with replacement, until each sample had the same number of sequences that it did in the observed data. Under the richness-adjusted community model, each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups. Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were generated for each dataset.

@@

where $x_{ij}$ was the number of sequences observed for OTU $i$ in sample $j$ and $g()$ was the geometric mean, $x_j$ was the count of the $S_T$ OTUs in sample $j$. Because the geometric mean is zero if any OTU is absent from a sample, the CLR is undefined when there are unobserved OTUs in a sample. To overcome this problem, I attempted a four approaches. The first, Zero CLR, imputed the value of the zeroes based on the observed data using the zCompositions (v.`r package_version("zCompositions")`) R package [@Palarea2015]. The second, One CLR, added a pseudo-count of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020]. The third, Nudge CLR, added a pseudo-count of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020]. The final approach, Robust CLR, removed unobserved OTUs prior to calculating the CLR [@Martino2019].

@@

**Table 1. Summary of studies used in the analysis.** For all studies, when rarefaction was used the number of sequences used from each dataset was the size of the smallest sample. A graphical representation of the distribution of sample sizes for each dataset and the samples that were removed from each dataset are provided in Figure S1.

@@

**Figure 1. Rarefaction eliminates the correlation between sequencing depth and alpha diversity (A) and between differences in squencing depth and beta (B) diversity metrics when using null community models.** Examples of the relationship between different metrics and methods for controlling for uneven sequencing effort are provided in Figures S2 and S3 for alpha and beta diversity metrics, respectively. Each point represents the mean of 100 random null community models; the standard deviation was smaller than the size of the plotting symbol.

@@

**Figure 2. The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5%, regardless of approach for controlling for uneven sequencing depth**. Samples were randomly assigned to different treatment groups. To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold. The false positive rate was the number of times a dataset yeilded a significant result.

@@

**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**. Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset. To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold. The false positive rate was the number of times a dataset yeilded a significant result.

@@

**Figure S1. The number of sequences observed in each sample for each dataset included in this analysis generally varied by 10 to 100-fold.** The threshold for specifying the number of sequences per sample varied by dataset and was determined based on identifying natural breaks in the data.

@@

**Figure S2. Examples of the richness in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the number of sequences in each of the samples prior to each method's approach of controlling for uneven sequencing effort. The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.

@@

**Figure S3. Examples of differences in beta diversity in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the difference in the number of sequences in each of the samples that went into calculating the pairwise distance prior to each method's approach of controlling for uneven sequencing effort. The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.

>>>>

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet}
 - \renewcommand*\familydefault{\sfdefault}
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

\raggedright

@@

${\dagger}$ To whom correspondence should be addressed:

@@

Considering it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences found across samples in a study, researchers need to control for the effect of uneven sequencing effort. How to do this has become a contentious question. Some have argued that rarefying or rarefaction is "inadmissible" because it omits valid data. A number of alternative approaches have been developed to normalize and rescale the data that purport to be invariant to the number of observations. I generated community distributions based on 12 published datasets where I was able to assess the ability of multiple methods to control for uneven sequencing effort. Rarefaction was the only method that could control for variation in uneven sequencing effort when measuring commonly used alpha and beta diversity metrics. Next, I compared the false detection rate and power to detect true differences between simulated communities with a known effect size using various alpha and beta diversity metrics. Although all methods of controlling for uneven sequencing effort had an acceptable false detection rate when samples were randomly assigned to two treatment groups, rarefaction was consistently able to control for differences in sequencing effort when sequencing depth was confounded with treatment group. Finally, the statistical power to detect differences in alpha and beta diversity metrics was consistently the highest when using rarefaction. These simulations underscore the importance of using rarefaction to normalize the number of sequences across samples in amplicon sequencing analyses.

@@

The ability to generate millions of 16S rRNA gene amplicon and metagenomic sequence reads has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that are embedded into the sequence of the PCR primers [@Sogin2006; @Caporaso2010; @Kozich2013]. Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1). This occurs because pooling of DNA from multiple PCRs is fraught with numerous opportunities for technical errors to compound leading to a skewed distribution. Aside from developing better methods of pooling DNA, the question of how to control for uneven sequencing effort in microbial ecology studies has become controversial.

The practice of rarefaction has been commonly used in ecology for more than 50 years as a tool to control for uneven sequencing effort across experimental replicates [@Sanders1968; @Hurlbert1971]. Microbial ecologists have used it to compare 16S rRNA gene sequence data for the past 25 years [@Dunbar1999; @Hughes2001; @Moyer1998]. With rarefaction the investigator selects a desired threshold of sequencing effort and removes any samples below that threshold. They then randomly select that many sequences, without replacement from each sample. Based on the observed sequence counts, the researcher can then calculate alpha diversity metrics including richness and diversity indices or beta diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index. I refer to this single sampling as a subsample; this method is implemented as the `sub.sample` function in mothur [@Schloss2009] and the `rrarefy` function in the `vegan` R package [@Dixon2003]. Rarefaction repeats the subsampling a large number of times (e.g., 100 or 1,000 times) and calculates the mean of the alpha or beta diversity metric over those subsamplings; rarefaction is implemented in mothur using the `summary.single` and `dist.shared` functions [@Schloss2009] and with the `vegan` R package using the `rarefy` or `avgdist` functions [@Dixon2003]. Rarefaction effectively tells a researcher what an alpha or beta diversity metric would have been for a collection of samples if they were all sequenced to the same depth. Although a closed form equation exists to calculate the expected richness [@Hurlbert1971], it is computationally easier to empirically calculate richness and other alpha and beta diversity metrics by rarefaction.

In 2014, McMurdie and Holmes [@McMurdie2014] announced that "rarefying" of microbial community data was "statistically inadmissible" because it omits valid data. In their simulations, they observed that rarefying reduced the statistical power to correctly cluster samples into the same treatment groups based on beta diversity metrics. Although the detail was lost on many subsequent researchers, they did not describe rarefaction, but rarefying [@Schloss2023wnwn]. According to McMurdie and Holmes, rarefying is a single subsampling of an OTU abundance table whereas rarefaction repeats the subsampling step many times. Furthermore, their simulations penalized the rarefied data by removing 15% of the samples and evaluated clustering quality using a clustering algorithm that performed worse for rarefied data. According to my reanalysis of rarefaction using the original simulation framework, rarefaction outperformed other normalization method being used in 2014 for both alpha and beta diversity metrics [@Schloss2023wnwn]. Others have also critiqued the original work [@Weiss2017]. Regardless, there is a general sense in the community that rarefying and rarefaction should not be used with microbiome data.

Since McMurdie and Holmes published their simulations, others have developed alternative approaches to control for uneven sequencing effort in amplicon sequencing studies. For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019]. Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [@Willis2015] have been used in microbial ecology studies. For beta diversity metrics at least four approaches have been pursued. First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020]. Second, normalization strategies have been developed where the relative abundance is multiplied by the size of the minimum desired sequencing effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013]. Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014]. This strategy is purported to control for uneven sequencing effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021]. Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].

The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the McMurdie and Holmes study. My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1). The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects. For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sequencing effort. My overall conclusion was that rarefaction outperformed the alternative strategies.

@@

***Without rarefaction, metrics of beta diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring beta diversity to sequencing effort, I used the same null models used for studying the sensitivity of alpha diversity. Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3). First, I calculated the Jaccard distance coefficient between all pairs of communities within a dataset. The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs. Jaccard distances were calculated using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS. Only the distances calculated with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU. Similar to what I observed with the Jaccard coefficient, only the data normalized with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix completion to replace zero counts (Robust CLR), adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR). The Aitchison distances were all strongly sensitive to sequencing effort (Figure 1B). Finally, I used the variance-stabilization transformation (VST) from DESeq2 prior to calculating Euclidean distances. Again, there was a strong sensitivity to sequencing effort (Figure 1B). Although Euclidean distances are not typically used on raw or rarefaction-normalized count data in ecology [@Legendre2012], Euclidean distances calculated with rarefaction were not sensitive to sequencing effort. Across each of the beta diversity metrics and approaches used to account for uneven sequencing effort and sparsity, rarefaction was the least sensitive approach to differences in sequencing effort.

@@

```{r coverage}
loss <- table_1 %>%
  mutate(min_max = 100 * min / max) %>%
  select(nice_name, min_max) %>%
  slice_min(min_max)

@@

***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$min_max`% of sequences that were in the largest sample). To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6). The richness values increased with sequencing effort as rare OTUs would continue be detected. In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sequencing effort. This result was expected since increased sequencing would lead to increased precision in the measured relative abundance of the OTUs. Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances. Although the mean richness appeared to increased unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable. In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths. Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase. Regardless, the coefficients of variation were relatively small. This analysis indicates that there are benefits to increased sequencing depths.

@@

All simulations have weaknesses and should be interpreted with caution. However, the simulated communities generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups. Indeed, is is perfectly admissible and proper to use rarefaction with microbial community data. The alternative is to risk reaching unwarranted conclusions.

@@

**Null treatment models.** I created an unbiased and biased treatment model. In the unbiased model, samples were randomly assigned to one of two treatment groups. In the biased treatment model, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group. Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\alpha$) of 0.05.

@@

**Richness-adjusted community model.** In the richness-adjusted community model, communities were randomly assigned to one of two simulated treatment groups. Communities in the first treatment group were generated by calculating the relative abundance of each OTU across all samples and using those values as the probability of sampling each OTU. This probability distribution was sampled until each sample had the same number of sequences that it did in the observed data. Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs. Sequences were sampled from the richness-adjusted community distribution, with replacement, until each sample had the same number of sequences that it did in the observed data. Under the richness-adjusted community model, each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups. Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were generated for each dataset.

@@

where $x_{ij}$ was the number of sequences observed for OTU $i$ in sample $j$ and $g()$ was the geometric mean, $x_j$ was the count of the $S_T$ OTUs in sample $j$. Because the geometric mean is zero if any OTU is absent from a sample, the CLR is undefined when there are unobserved OTUs in a sample. To overcome this problem, I attempted a four approaches. The first, Zero CLR, imputed the value of the zeroes based on the observed data using the zCompositions (v.`r package_version("zCompositions")`) R package [@Palarea2015]. The second, One CLR, added a pseudo-count of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020]. The third, Nudge CLR, added a pseudo-count of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020]. The final approach, Robust CLR, used matrix completion to replace zero counts for unobserved OTUs prior to calculating the CLR [@Martino2019].

@@

**Table 1. Summary of studies used in the analysis.** For all studies, when rarefaction was used the number of sequences used from each dataset was the size of the smallest sample. A graphical representation of the distribution of sample sizes for each dataset and the samples that were removed from each dataset are provided in Figure S1. This table is similar to Table 1 from [@Schloss2023sing].

@@

**Figure 1. Rarefaction eliminates the correlation between sequencing depth and alpha diversity (A) and between differences in sequencing depth and beta (B) diversity metrics when using null community models.** Examples of the relationship between different metrics and methods for controlling for uneven sequencing effort are provided in Figures S2 and S3 for alpha and beta diversity metrics, respectively. Each point represents the mean of 100 random null community models; the standard deviation was smaller than the size of the plotting symbol.

@@

**Figure 2. The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5%, regardless of approach for controlling for uneven sequencing depth**. Samples were randomly assigned to different treatment groups. To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold. The false positive rate was the number of times a dataset yielded a significant result.

@@

**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**. Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset. To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold. The false positive rate was the number of times a dataset yielded a significant result.

@@

**Figure S1. The number of sequences observed in each sample for each dataset included in this analysis generally varied by 10 to 100-fold.** The threshold for specifying the number of sequences per sample varied by dataset and was determined based on identifying natural breaks in the data. This figure is similar to Figure S1 of [@Schloss2023sing]

@@

**Figure S2. Examples of the richness in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the number of sequences in each of the samples prior to each method's approach of controlling for uneven sequencing effort. The Spearman correlation coefficient ($\rho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.

@@

**Figure S3. Examples of differences in beta diversity in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the difference in the number of sequences in each of the samples that went into calculating the pairwise distance prior to each method's approach of controlling for uneven sequencing effort. The Spearman correlation coefficient ($\rho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
